Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2022-02-07 _paper.rmd
house prices and conclusion

(ref:casepenguins) Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is an Chinstrap (orange) penguin that is misclassified as a Gentoo (purple), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of `f_l` from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at [vimeo.com/666431172](https://vimeo.com/666431172).)
```{r casepenguins, echo=F, out.width="100%", fig.cap = "(ref:casepenguins)"}
knitr::include_graphics("./figures/case_penguins.png")
```

@@

(ref:caseames) Ames housing 2018 regressing sales price [USD]. The PI sale price was under predicted and had sizable attribution to lot area (`LtA`). The CI was predicted sales price was similar and much more accurate with its observed sales price while it has very little attribution to lot area. Varying the contribution lot area the separation between these house sales crosses when there is a low contribution of `LtA`, which is important to explaining the PI and near invariant to the sales price of the CI. (A video showing the animation is at [vimeo.com/666431134](https://vimeo.com/666431134).)
```{r caseames, echo=F, out.width="100%", fig.cap = "(ref:caseames)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

@@

The need to maintain the interpretability of black-box models is evident. One aspect uses local explanations of the model in the vicinity of an instance. Local explanations approximate the linear feature importance to the model. Our contribution is to assess explanations by examining the support by varying the contributions with a radial tour. First, a global view visualizes approximations of the data space, explanation space, model predictions side-by-side, using dynamic interaction to compare and contrast and identify instances of interest. The normalized linear importance from the explanation of the PI becomes the feature of interest to further explore with the radial tour. The tours explore the feature sensitivity to the structure identified in the explanation.

We have illustrated this method on random forest models using the tree SHAP local explanation. While this analysis could be generally used with any compatible model-explanation pairing. We have created an open-source __R__ package __cheem__, available on [CRAN](https://CRAN.R-project.org/package=cheem), to facilitate preprocessing and exploration with the described interactive application. Toy and real data are provided, or upload your data after preprocessing.

One limitation of the __cheem__ packages is that it currently only calculates tree SHAP for tree-based models supported by __treeshap__ (tree based models from __randomForest__, __ranger__, __gbm__, __xgboost__, __lightgbm__, or __catboost__). The SHAP and oscillation explainations could be added with use of the `DALEX::explain()` seems to be an excellent direction to extend [@biecek_dalex_2018; @biecek_explanatory_2021]. Tree SHAP was selected because of its reduced computational complexity; processing runtime of the explanation will be a looming concern.

Attribution space does tend to be more separable than data space at least in the first couple principal components. Different statistics should be explored to convey a clearer understanding what structure the explanation distinguishes. It is curious that the attribution space the same dimensionality of the data with a seemingly better separation. It sounds like a candidate to fit another model. However, such an attempt should want to be vigilant not to over-fit the training data.

>>>>

(ref:casepenguins) Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is a Gentoo (purple) Chinstrap (orange) penguin that is misclassified as a Chinstrap (orange), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of `f_l` from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably (higher density value) a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at [vimeo.com/666431172](https://vimeo.com/666431172).)
```{r casepenguins, echo=F, out.width="100%", fig.cap = "(ref:casepenguins)"}
knitr::include_graphics("./figures/case_penguins.png")
```

@@

(ref:caseames) Exploring an instance with a large residual as the PI from fitting sales price [USD] to other variables in the Ames housing 2018. The sale price of the PI was under-predicted. The local explanation indicates a sizable attribution to lot area (`LtA`). The CI has a similar predicted sales price and smaller residual, and has very little attribution to lot area. In the attribution projection, the PI has a higher sales price than the CI. Reducing the contribution of lot area brings these two prices in line. This suggests if lot area was not considered by the model, then the two houses would be quite similar. That is, the large residual is due to a lack of factoring in the lot area for the prediction of PI's sales price. (A video showing the animation is at [vimeo.com/666431134](https://vimeo.com/666431134).)
```{r caseames, echo=F, out.width="100%", fig.cap = "(ref:caseames)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

@@

New tools to assist with the interpretability of black-box models is increasingly important. This paper provides a technique that builds on local interpretations to explore the feature importance local to an instance. The local interpretations form an attribution projection from which feature contributions are varied using a radial tour. Several diagnostic plots are provided to assist with understanding the sensitivity of the prediction to particular features. A global view shows the data space, explanation space, model fit. The user can interactively select instances to compare and contrast, and study further.

The approach has been illustrated on four data examples using random forest models and using the tree SHAP local explanation. In the penguins example, we showed how the misclassification of a penguin arose due to it having an unusually small flipper size compared to others of it's species. This was verified by making a follow-up plot of the data including this variable. In the chocolates example, we showed how a dark chocolate was misclassified primarily due to it value on Fiber, and a milk chocolate misclassified as dark due to it's lowish Sodium value. For the FIFA example, we show how low Messi's salary would go if it depended on defensive skills. In the Ames housing data, an inaccurate prediction for a house was likely due to lot area not being effectively used.

The approach is manually intensive, and thus only feasible for investigating a few instances. The approach is to select instances which the model hasn't done well and compare it with an instance where the model did fit well.  The radial tour launches from the attribution projection, to enable exploration of the sensitivity of the prediction to any feature. It can be helpful to make additional plots of the features and response in order to cross-check interpretations made from the cheem viewer. This methodology provides an additional tool in the box for studying model fitting.

An implemenation is provided in the open-source __R__ package __cheem__, available on [CRAN](https://CRAN.R-project.org/package=cheem). Example data sets are provided, and you can upload your own data after model fitting and computing the local explanaitons. In theory this approach would work with any black box model but the implementation currently only calculates tree SHAP for tree-based models supported by __treeshap__ (tree based models from __randomForest__, __ranger__, __gbm__, __xgboost__, __lightgbm__, or __catboost__). Tree SHAP was selected because of its computational efficiency. The SHAP and oscillation explanations could be added with use of the `DALEX::explain()` and would be an excellent direction to extend the work [@biecek_dalex_2018; @biecek_explanatory_2021].

<!-- Attribution space does tend to be more separable than data space at least in the first couple principal components. Different statistics should be explored to convey a clearer understanding what structure the explanation distinguishes. It is curious that the attribution space the same dimensionality of the data with a seemingly better separation. It sounds like a candidate to fit another model. However, such an attempt should want to be vigilant not to over-fit the training data.-->

### 2022-02-07 _paper.tex
house prices and conclusion

\caption{Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is an Chinstrap (orange) penguin that is misclassified as a Gentoo (purple), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of \texttt{f\_l} from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at \href{https://vimeo.com/666431172}{vimeo.com/666431172}.)}\label{fig:casepenguins}
\end{figure}

@@

\caption{Ames housing 2018 regressing sales price {[}USD{]}. The PI sale price was under predicted and had sizable attribution to lot area (\texttt{LtA}). The CI was predicted sales price was similar and much more accurate with its observed sales price while it has very little attribution to lot area. Varying the contribution lot area the separation between these house sales crosses when there is a low contribution of \texttt{LtA}, which is important to explaining the PI and near invariant to the sales price of the CI. (A video showing the animation is at \href{https://vimeo.com/666431134}{vimeo.com/666431134}.)}\label{fig:caseames}
\end{figure}

@@

The need to maintain the interpretability of black-box models is evident. One aspect uses local explanations of the model in the vicinity of an instance. Local explanations approximate the linear feature importance to the model. Our contribution is to assess explanations by examining the support by varying the contributions with a radial tour. First, a global view visualizes approximations of the data space, explanation space, model predictions side-by-side, using dynamic interaction to compare and contrast and identify instances of interest. The normalized linear importance from the explanation of the PI becomes the feature of interest to further explore with the radial tour. The tours explore the feature sensitivity to the structure identified in the explanation.

We have illustrated this method on random forest models using the tree SHAP local explanation. While this analysis could be generally used with any compatible model-explanation pairing. We have created an open-source \textbf{R} package \textbf{cheem}, available on \href{https://CRAN.R-project.org/package=cheem}{CRAN}, to facilitate preprocessing and exploration with the described interactive application. Toy and real data are provided, or upload your data after preprocessing.

One limitation of the \textbf{cheem} packages is that it currently only calculates tree SHAP for tree-based models supported by \textbf{treeshap} (tree based models from \textbf{randomForest}, \textbf{ranger}, \textbf{gbm}, \textbf{xgboost}, \textbf{lightgbm}, or \textbf{catboost}). The SHAP and oscillation explainations could be added with use of the \texttt{DALEX::explain()} seems to be an excellent direction to extend (Biecek 2018; Biecek and Burzykowski 2021). Tree SHAP was selected because of its reduced computational complexity; processing runtime of the explanation will be a looming concern.

Attribution space does tend to be more separable than data space at least in the first couple principal components. Different statistics should be explored to convey a clearer understanding what structure the explanation distinguishes. It is curious that the attribution space the same dimensionality of the data with a seemingly better separation. It sounds like a candidate to fit another model. However, such an attempt should want to be vigilant not to over-fit the training data.

>>>>

\caption{Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is a Gentoo (purple) Chinstrap (orange) penguin that is misclassified as a Chinstrap (orange), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of \texttt{f\_l} from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably (higher density value) a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at \href{https://vimeo.com/666431172}{vimeo.com/666431172}.)}\label{fig:casepenguins}
\end{figure}

@@

\caption{Exploring an instance with a large residual as the PI from fitting sales price {[}USD{]} to other variables in the Ames housing 2018. The sale price of the PI was under-predicted. The local explanation indicates a sizable attribution to lot area (\texttt{LtA}). The CI has a similar predicted sales price and smaller residual, and has very little attribution to lot area. In the attribution projection, the PI has a higher sales price than the CI. Reducing the contribution of lot area brings these two prices in line. This suggests if lot area was not considered by the model, then the two houses would be quite similar. That is, the large residual is due to a lack of factoring in the lot area for the prediction of PI's sales price. (A video showing the animation is at \href{https://vimeo.com/666431134}{vimeo.com/666431134}.)}\label{fig:caseames}
\end{figure}

@@

New tools to assist with the interpretability of black-box models is increasingly important. This paper provides a technique that builds on local interpretations to explore the feature importance local to an instance. The local interpretations form an attribution projection from which feature contributions are varied using a radial tour. Several diagnostic plots are provided to assist with understanding the sensitivity of the prediction to particular features. A global view shows the data space, explanation space, model fit. The user can interactively select instances to compare and contrast, and study further.

The approach has been illustrated on four data examples using random forest models and using the tree SHAP local explanation. In the penguins example, we showed how the misclassification of a penguin arose due to it having an unusually small flipper size compared to others of it's species. This was verified by making a follow-up plot of the data including this variable. In the chocolates example, we showed how a dark chocolate was misclassified primarily due to it value on Fiber, and a milk chocolate misclassified as dark due to it's lowish Sodium value. For the FIFA example, we show how low Messi's salary would go if it depended on defensive skills. In the Ames housing data, an inaccurate prediction for a house was likely due to lot area not being effectively used.

The approach is manually intensive, and thus only feasible for investigating a few instances. The approach is to select instances which the model hasn't done well and compare it with an instance where the model did fit well. The radial tour launches from the attribution projection, to enable exploration of the sensitivity of the prediction to any feature. It can be helpful to make additional plots of the features and response in order to cross-check interpretations made from the cheem viewer. This methodology provides an additional tool in the box for studying model fitting.

An implemenation is provided in the open-source \textbf{R} package \textbf{cheem}, available on \href{https://CRAN.R-project.org/package=cheem}{CRAN}. Example data sets are provided, and you can upload your own data after model fitting and computing the local explanaitons. In theory this approach would work with any black box model but the implementation currently only calculates tree SHAP for tree-based models supported by \textbf{treeshap} (tree based models from \textbf{randomForest}, \textbf{ranger}, \textbf{gbm}, \textbf{xgboost}, \textbf{lightgbm}, or \textbf{catboost}). Tree SHAP was selected because of its computational efficiency. The SHAP and oscillation explanations could be added with use of the \texttt{DALEX::explain()} and would be an excellent direction to extend the work (Biecek 2018; Biecek and Burzykowski 2021).
